---
title: "Clustering concordance indexes for fuzzy partition based on Rand index "
author: "Hugo Barbot"
date: "`r Sys.Date()`"
bibliography:
  - biblio_fuzzy_partition.bib
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    theme: flatly
  # prettydoc::html_pretty:
  #   self_contained: true
  #   theme: cayman
  #   highlight: github
  #   toc: true
  #   css: styles.css  # sol de GPT, ne marche pas du premier coup, je verrai plus tard
  #   # toc_float: true
  #   # toc_depth: 3
  #   number_sections: true
  #   keep_tex: yes
---

```{=html}
<style>
body {text-align: justify}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", echo = TRUE, results="verbatim", fig.align = 'center')
```


# Rand, Adjusted Rand and Jaccard indexes


Clustering is a widely known machine learning problem in which the goal is to determnine a finite set of categories to describe the structure of a dataset according to similarities among objects. Let a clustering $C = \{C_1, C_2, \dots, C_k\}$ of $k$ subsets $C_k$ of our dataset $D = \{d_1, d_2, \dots, d_N\}$, such that $C_1 \ \cup C_2 \ \cup \dots \ \cup C_k = D$. Such a collection is referred to as a *partition*, because it imposes to the dataset a partitioning of its object into subsets (clusters/components). When all subsets are mutually disjoint, i.e. $C_i \cap C_l = \varnothing, \ \forall i,l \in {1,\dots,k} \ (i \neq l)$, the partition is said to be of *hard* type. A hard partition of $N$ objects into a certain number $k$ of disjoint subsets (called clusters in the case of hard type partition) can be represented by a $k \times N$ *partition matrix* whose element $p_{ij}$ is either 1 if the $j^{th}$ object belongs to the $i^{th}$ cluster or 0 otherwise, i.e.:

$$P = [p_{ij}]_{k \times N}; \ \sum_{i=1}^k p_{ij} = 1 \quad \forall j \in \{1,\dots,N\}$$

A problem that is often of interest is that of compairing two partitions of a given dataset, possibly with different numbers ($k$ and $\nu$) of subsets. Formally, given two partition matrices of the same dataset, $R = [r_{ij}]_{k \times N}$ and $Q = [q_{ij}]_{k \times N}$, the underlying problem is to properly measure the degree of matching between the partitions these matrices describe. In principle, it is not a trivial problem even when dealing with hard partitions matrices, for the following main reasons [@campello_generalized_2010]:
1. the numbers of clusters$k$ and $\nu$  in respectively matrices $R$ and $Q$ are not necessarily the same,
2. row permutations in $R$ and $Q$ do not change the partitions these matrices describe.

A very appealing and stringent solution to the problem of comparing partitions is the use of so-called external validity criteria for hard clustering assessment. Such criteria are termed *external* because, in contrast to their *internal* or *relative* conterparts, they use external information (another or a reference partition) - besides the data themself - to evaluate the quality of a given clustering structure. Most external criteria, as the Rand index, is constructed from the following set of building blocks based on pairwise comparisons of data objects:
- a: No. of pairs of data objects belonging to the same cluster in $R$ and to the same cluster in $Q$,
- b: No. of pairs of data objects belonging to the same cluster in $R$ and to different cluster in $Q$,
- c: No. of pairs of data objects belonging to different cluster in $R$ and to the same cluster in $Q$,
- d: No. of pairs of data objects belonging to different cluster in $R$ and to different cluster in $Q$,

The Rand index is an *external evaluation measure* developed by [@rand1971objective] to compare hard partitions on a set of data. This index is given by 
$$\text{RI}(R,Q) = \frac{a+d}{a+b+c+d}$$
and clearly exhibits the following desirable properties:
1. RI $\in [0,1]$,
2. RI $= 0$ iff $Q$ is completly inconsistent, i.e. $a = d = 0$,
3. RI $= 1$ iff both partitions match, i.e. $b = c = 0$.

Unfortunatly the Rand index approaches its upper limit as the number of clusters increases. There are some others known problems with the Rand index [@meilua2007Randpb]:
- The expected value of the Rand index for two random partitions does not take a constant value,
- It presents high variability and it concentrates in a small interval close to 1,
- It is extremely sensitive to the number of groups, their size and also to the overall number of observations considered in each partitions.

To overcome these problems, [@hubert1985ARI] proposed an adjusted version of the Rand index assuming the generalized hypergeometrical distribution as a model of randomness. In other words, this adjusted version is equal to the normalized difference of the Rand index and its expected value under the null hypothesis:
$$\text{ARI}(R,Q) = \frac{\textit{Index} - \textit{Expected Index}}{\textit{Maximum Index} - \textit{Expected Index}} = \frac{2(ad - bc)}{b^2 + c^2 + 2ad + (a+d)(b+c)}.$$
This index has an upper bound of 1 and takes the value 0 when the Rand index is equal to its expected value. Negative value are possible and indicate less agreement between $R$ and $Q$ than expected by chance.


For example, lets have a reference partition $R$ and two partition, $Q_1$ and $Q_2$, with the same clustering but not the same cluster label, along with two partitions, $Q_3$ and $Q_4$, with one observation in a different cluster and with different cluster label. Those hard partitions are being vectorise, but can be easily transformed in matrix as defined earlier.

```{r data_ARI_example}
data <- structure(
  list(
    R = c(1, 1, 1, 2, 2, 2, 2, 3, 3, 3, 3, 3), 
    Q1 = c(3, 3, 3, 1, 1, 1, 1, 2, 2, 2, 2, 2), 
    Q2 = c(2, 2, 2, 3, 3, 3, 3, 1, 1, 1, 1, 1), 
    Q3 = c(1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 3, 3), 
    Q4 = c(3, 3, 3, 3, 1, 1, 1, 2, 2, 2, 2, 2)), 
  row.names = c(NA, -12L), 
  class = "data.frame")
```

We can use the adjusted Rand index:
```{r ARI_example}
mclust::adjustedRandIndex(data$R, data$R)
mclust::adjustedRandIndex(data$R, data$Q1)
mclust::adjustedRandIndex(data$R, data$Q2)
mclust::adjustedRandIndex(data$R, data$Q3)
mclust::adjustedRandIndex(data$R, data$Q4)
```

ARI is 1 in the first three conditions, the only difference being the cluster label. On the other hand, partitions $Q_3$ and $Q_4$ the same partitioning which is different from the previous one. So adjustedRandIndex(data$R, data$Q3) and adjustedRandIndex(data$x, data$Q4) is different from 1 but both are the same.



Another common criticism against both the Rand index and its adjusted version is that it gives the same importance to both the agreement terms $a$ and $d$, thus making no difference between pairs of objects that are joined or separated in both partitions $R$ and $Q$. Indeed it is known that the term $d$ may dominate the others three terms, thus causing the last problem stated above. The ordinary removal of term $d$ from the Rand index results in the so-called Jaccard coefficient:
$$\text{J}(R,Q) = \frac{a}{a+b+c}.$$

All these indexes, however, allow uniquely the evaluation of hard clustering partitions, at least as they were originally formulated. Some fully equivalent set-theoretic formulation for all of them was derived and extended to more complex partitions [@campello2007fuzzy]. However, it has the shortcoming that one of the partitions must be hard. The same author proposed a general approach for comparing two data partitions with overlapping categories based on optimal transport (*mais je la trouve pas hyper optimal*).




# Definition of more complexe partition and generalisation of previous indexes


*voir section 2 et 3 de [@campello_generalized_2010]*

*voir également section 3 de [@dambrosio_adjusted_2021]*



# Normalized degree of concordance(NDC) and adjusted concordance index (ACI): extensions of the Rand index to fuzzy partition

*voir section de [@dambrosio_adjusted_2021]*

